United Monoids

### 2022-02-05 paper/main.tex
More tweaks

% Fix broken footnotes
\makeatletter
\renewcommand\@makefntext[1]{\leftskip=0.7em\hskip-0.75em\@makefnmark\hspace{1mm}#1}
\makeatother
\renewcommand{\footnotesize}{\small}
\setlength{\footnotesep}{3mm}

@@

\newcommand{\code}[1]{\lstinline[mathescape]|#1|}
\newcommand{\hcode}[1]{{\color{darkblue} \lstinline[keywordstyle={}]|#1|}} % h for "highlighted"
\newcommand{\h}[1]{{\itshape\color{grayblue}#1}} % h for "highlighted"
\newcommand{\n}[1]{{\itshape\color{graygreen}#1}} % n for "normal"
\newcommand{\up}[1]{{\itshape\color{graygreen}\textsuperscript{#1}}}
\newcommand{\hadd}{{\large\color{darkblue} $\oplus$}}
\newcommand{\hmul}{{\large\color{darkblue} $\otimes$}}
\newcommand{\hdia}{\,\text{\raisebox{-0.2mm}{\Large\color{darkblue} $\diamond$}}\,}
\newcommand{\harr}{\,\text{\color{darkblue} $\rightarrow$}\,}

@@

\title{United Monoids}
\subtitle{Finding Simplicial Sets and Labelled Algebraic Graphs in Trees}
\author[~]{Andrey Mokhov}
\authorinfo{is a software engineer at Jane Street London,
  and a visiting fellow at Newcastle University, UK. His
  research interests are in applying abstract mathematics and
  functional programming to solving large-scale engineering
  problems.
  \begin{description}
    \item[email:] \email{andrey.mokhov@ncl.ac.uk}
  \end{description}}
\affiliation[~]{\hspace{-5.3mm}Jane Street, London, UK}
\affiliation[~]{\hspace{-5.3mm}Newcastle University, Newcastle upon Tyne, UK}

@@

\begin{lstlisting}[float,label=lst-tree-std,xleftmargin=0pt,caption={
    The \code{Tree} data type and instances of various standard Haskell type classes.
}]
!\hrulefill!
data Tree s a = Leaf a | Node s (Tree s a) (Tree s a)

@@

instance IsString a => IsString (Tree s a) where -- This makes !\h{"a"}! a shortcut for !\h{Leaf "a"}!
    fromString = Leaf . fromString
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

\begin{lstlisting}[float,label=lst-tree,xleftmargin=0pt,belowskip=-2mm,caption={
    A basic API for constructing and manipulating trees. We use the standard
    module \code{Data.L}\code{ist.N}\code{onEmpty} for representing non-empty
    lists, and \code{Data.S}\code{et} for sets.}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
leaf :: a -> Tree s a -- A tree with a single leaf
leaf = Leaf

@@

filter :: (a -> Bool) -> Tree s a -> Maybe (Tree s a) -- Prune all leaves !\h{a}! with !\h{p a = False}!
filter p = traverse (\a -> if p a then Just a else Nothing) -- Via !\h{Traversable}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

\section{Sets}\label{sec-set}

@@

\begin{lstlisting}[label=lst-set,xleftmargin=0pt,belowskip=7pt,caption={
    Implementing a part of the standard \code{Data.S}\code{et} API with
    \hcode{TSet}.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TSet a = Tree () a -- Sets are trees with !\h{()}!-labelled internal nodes

@@

map :: (a -> b) -> TSet a -> TSet b
map = fmap -- Via !\h{Functor}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

In addition to associativity, the set interpretation of trees puts two new
requirements on every call site of \hcode{fold leaf node}\footnote{
    Haskell's type system isn't powerful enough to check such requirements, so
    we should use \hcode{fold} with care. A dependently typed language, such as
    Agda~\cite{2007_norell_agda}, could help us here.
}:

@@

\begin{lstlisting}[float,label=lst-simplicial-set,xleftmargin=0pt,caption={
    A basic API for working with simplicial sets represented by trees.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TSimplicialSet a = Tree Bool a -- Simplicial sets are trees over the Boolean semiring

@@

map :: (a -> b) -> TSimplicialSet a -> TSimplicialSet b -- The so-called !\n{``}!simplicial map!\n{''}!
map = fmap -- Via !\h{Functor}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

\begin{lstlisting}[float,label=lst-graph,xleftmargin=0pt,caption={
    Implementing a part of the \code{Algebra.G}\code{raph} API~\cite{mokhov_alga}
    with \hcode{Tree Bool a}.
}]
!\rule[2mm]{\textwidth}{0.4pt}!
!\vspace{-7mm}!
type TGraph a = Tree Bool a -- Graphs are trees over the Boolean semiring

@@

hasEdge :: Ord a => a -> a -> TGraph a -> Bool
hasEdge x y graph = case induce (\a -> a == x || a == y) graph of
    Nothing -> False
    Just sub -> elem (x,!\,!y) (edgeSet sub) -- Here !\h{sub}! can only contain vertices !\h{x}! and !\h{y}!
!\rule[2mm]{\textwidth}{0.4pt}!
\end{lstlisting}

@@

\noindent
When implementing \hcode{connect}, we come across a subtle issue with
\hcode{LGraph}: it is possible to unintentionally accumulate redundant
\zero-labelled edges in the inner \hcode{Map}. Filtering out these redundant
edges from \hcode{LGraph s a} is an important optimisation when working
with sparse graphs in practice, which is why the code below requires
\hcode{Eq s}\footnote{
    We do not need the full power of \hcode{Eq s} here: we just need to test if
    a given edge label is~\zero. For some types \hcode{s}, testing for \zero is
    much easier than a general equality test; for example, testing if a
    \hcode{Map k v} is empty is trivial but \hcode{Eq (Map k v)} is undecidable
    for some choices of \hcode{v}. To resolve this problem, one can switch to a
    more precise constraint like \hcode{EqZero s}.
}. Note that the \hcode{Tree}-based representation of graphs sidesteps this
issue because \hcode{overlay x y} and \hcode{connect zero x y} are represented
by the same tree \hcode{Node zero x y}.

@@

\begin{lstlisting}[basicstyle=\sffamily\smaller]
data Graph a = Empty | Vertex a | Overlay (Graph a) (Graph a) | Connect (Graph a) (Graph a)
\end{lstlisting}

>>>>

% Fix broken footnotes
\makeatletter
\renewcommand\@makefntext[1]{\leftskip=0.7em\hskip-0.75em\@makefnmark\hspace{1mm}#1}
\makeatother
% \renewcommand{\footnotesize}{\small}
% \setlength{\footnotesep}{3mm}

@@

\newcommand{\code}[1]{\lstinline[mathescape]|#1|}
\newcommand{\hcode}[1]{{\color{darkblue} \lstinline[keywordstyle={}]|#1|}} % h for "highlighted"
\newcommand{\hfcode}[1]{{\color{darkblue} \lstinline[keywordstyle={},basicstyle=\footnotesize\ttfamily]|#1|}} % h for "highlighted"
\newcommand{\h}[1]{{\itshape\color{grayblue}#1}} % h for "highlighted"
\newcommand{\n}[1]{{\itshape\color{graygreen}#1}} % n for "normal"
\newcommand{\up}[1]{{\itshape\color{graygreen}\textsuperscript{#1}}}
\newcommand{\hadd}{{\large\color{darkblue} $\oplus$}}
\newcommand{\hmul}{{\large\color{darkblue} $\otimes$}}
\newcommand{\hdia}{\,\text{\raisebox{-0.2mm}{\Large\color{darkblue} $\diamond$}}\,}
\newcommand{\harr}{\,\text{\color{darkblue} $\rightarrow$}\,}

@@

\title{United Monoids}
\subtitle{Finding Simplicial Sets and Labelled Algebraic Graphs in Trees}
\author[\hspace{0.3mm}a,b]{Andrey Mokhov}
\authorinfo{is a software engineer at Jane Street London,
  and a visiting fellow at Newcastle University, UK. His
  research interests are in applying abstract mathematics and
  functional programming to solving large-scale engineering
  problems.
  \begin{description}
    \item[email:] \email{andrey.mokhov@ncl.ac.uk}
  \end{description}}
\affiliation[a]{\hspace{-1mm}Jane Street, London, UK}
\affiliation[b]{\hspace{-1mm}Newcastle University, Newcastle upon Tyne, UK}

@@

\begin{lstlisting}[float,label=lst-tree-std,xleftmargin=0pt,caption={
    The \code{Tree} data type and instances of various standard Haskell type classes.
},captionpos=t]
!\vspace{0mm}!
data Tree s a = Leaf a | Node s (Tree s a) (Tree s a)

@@

instance IsString a => IsString (Tree s a) where -- This makes !\h{"a"}! a shortcut for !\h{Leaf "a"}!
    fromString = Leaf . fromString
\end{lstlisting}

\begin{lstlisting}[float,label=lst-tree,xleftmargin=0pt,belowskip=-2mm,caption={
    A basic API for constructing and manipulating trees. We use the standard
    module \code{Data.L}\code{ist.N}\code{onEmpty} for representing non-empty
    lists, and \code{Data.S}\code{et} for sets.},captionpos=t]
!\vspace{-2mm}!
leaf :: a -> Tree s a -- A tree with a single leaf
leaf = Leaf

@@

filter :: (a -> Bool) -> Tree s a -> Maybe (Tree s a) -- Prune all leaves !\h{a}! with !\h{p a = False}!
filter p = traverse (\a -> if p a then Just a else Nothing) -- Via !\h{Traversable}!
!\vspace{-2mm}!
\end{lstlisting}

@@

\vspace{-1mm}
\section{Sets}\label{sec-set}
\vspace{-0.5mm}

@@

\begin{lstlisting}[label=lst-set,xleftmargin=0pt,belowskip=7pt,caption={
    Implementing a part of the standard \code{Data.S}\code{et} API with
    \hcode{TSet}.
},captionpos=t]
!\vspace{-1mm}!
type TSet a = Tree () a -- Sets are trees with !\h{()}!-labelled internal nodes

@@

map :: (a -> b) -> TSet a -> TSet b
map = fmap -- Via !\h{Functor}!
!\vspace{0mm}!
\end{lstlisting}

@@

In addition to associativity, the set interpretation of trees puts two new
requirements on every call site of \hcode{fold leaf node}\footnote{
    Haskell's type system isn't powerful enough to check such requirements, so
    we should use \hfcode{fold} with care. A dependently typed language, such as
    Agda~\cite{2007_norell_agda}, could help us here.
}:

@@

\begin{lstlisting}[float,label=lst-simplicial-set,xleftmargin=0pt,caption={
    A basic API for working with simplicial sets represented by trees.
},captionpos=t]
!\vspace{0mm}!
type TSimplicialSet a = Tree Bool a -- Simplicial sets are trees over the Boolean semiring

@@

map :: (a -> b) -> TSimplicialSet a -> TSimplicialSet b -- The so-called !\n{``}!simplicial map!\n{''}!
map = fmap -- Via !\h{Functor}!
\end{lstlisting}

@@

\begin{lstlisting}[float,label=lst-graph,xleftmargin=0pt,caption={
    Implementing a part of the \code{Algebra.G}\code{raph} API~\cite{mokhov_alga}
    with \hcode{Tree Bool a}.
},captionpos=t]
!\vspace{0mm}!
type TGraph a = Tree Bool a -- Graphs are trees over the Boolean semiring

@@

hasEdge :: Ord a => a -> a -> TGraph a -> Bool
hasEdge x y graph = case induce (\a -> a == x || a == y) graph of
    Nothing -> False
    Just sub -> elem (x,!\,!y) (edgeSet sub) -- Here !\h{sub}! can only contain vertices !\h{x}! and !\h{y}!
\end{lstlisting}

@@

\noindent
When implementing \hcode{connect}, we come across a subtle issue with
\hcode{LGraph}: it is possible to unintentionally accumulate redundant
\zero-labelled edges in the inner \hcode{Map}. Filtering out these redundant
edges from \hcode{LGraph s a} is an important optimisation when working
with sparse graphs in practice, which is why the code below requires
\hcode{Eq s}\footnote{
    We do not need the full power of \hfcode{Eq s} here: we just need to test if
    a given edge label is~\zero. For some types \hfcode{s}, testing for \zero is
    much easier than a general equality test; for example, testing if a
    \hfcode{Map k v} is empty is trivial but \hfcode{Eq (Map k v)} is undecidable
    for some choices of \hfcode{v}. To resolve this problem, one can switch to a
    more precise constraint like \hfcode{EqZero s}.
}. Note that the \hcode{Tree}-based representation of graphs sidesteps this
issue because \hcode{overlay x y} and \hcode{connect zero x y} are represented
by the same tree \hcode{Node zero x y}.

@@

\begin{lstlisting}[basicstyle=\sffamily\small]
data Graph a = Empty | Vertex a | Overlay (Graph a) (Graph a) | Connect (Graph a) (Graph a)
\end{lstlisting}
